quollr: An R Package for Visalizing 2D Models in High Dimensional Space

### 2024-02-26 paper-quollr.Rmd
Add the hexagonal binning logic with images

```{r datasets-tb-pdf, eval=is_latex_output(), echo=FALSE}
datasets_tb |>
  kable(caption = "quollr datasets", format="latex", col.names = c("data", "explanation"), booktabs = T)  |>
  column_spec(1, width = "4cm") |>
  column_spec(2, width = "8cm")
```

### Compute hexagonal bin configurations

@@

Before computing hexagonal bin configurations, we need to determine the range of data along the x and y axes to establish the boundary for hexagonal binning. Additionally, choosing an appropriate hexagon size (the radius of the outer circle) is essential. By default, the `calculate_effective_x_bins()` and `calculate_effective_y_bins()` functions use a hexagon size of 1.07457. However, users can adjust the hexagon size to fit the data range and achieve regular hexagons without overlapping.

@@

@@

@@

Generating full hexagonal grid contains main three steps:

@@

@@

@@

5. Assign data into hexagons

@@

6. Compute standardized counts

@@

7. Extract full grid info

- Assign standardize counts for hex bins

- Join with the hexagonal coordinates

@@

#### Buffer size

@@

Overall, including a buffer when generating hexagonal bins helps to produce more accurate and robust binning results, particularly when dealing with real-world data that may have irregular distributions or boundary effects.

### Construct the 2D model with different options

@@

#### Shift the hexagonal grid origin

If shift_x happen to the positive direction of x it should input as a positive value, if not other way
If shift_y happen to the positive direction of y it should input as a positive value, if not other way

1. Assign shift along the x and y axis (limited the amount should less than the cell_diameter)

2. Generate bounds with shift origin

>>>>

```{r datasets-tb-pdf, eval=is_latex_output(), echo=FALSE}
datasets_tb |>
  kable(caption = "quollr datasets", format="latex", col.names = c("data", "explanation"), booktabs = T)  |>
  column_spec(1, width = "4cm") |>
  column_spec(2, width = "8cm")
```
## Preprocessing

In the preprocessing stage, we aim to prepare our data to fit within the bounds required for regular hexagonal binning, ensuring effective visualization. To achieve this, we implement two key scaling steps. Firstly, we scale the first 2D embedding component to range between $0$ and $1$, ensuring that all data points fall within this normalized interval. Secondly, we scale the second 2D embedding component to range between $-\frac{\sqrt{3}}{2}$ and $\frac{\sqrt{3}}{2}$. This adjustment helps maintain the necessary symmetry and spacing required for regular hexagons.

## Compute hexagonal bin configurations

@@

Before computing hexagonal bin configurations, we need to determine the range of data along the x and y axes to establish the boundary for hexagonal binning. Additionally, choosing an appropriate hexagon size (the radius of the outer circle) is essential. By default, the `calculate_effective_x_bins()` and `calculate_effective_y_bins()` functions use a hexagon size of $0.2$. However, users can adjust the hexagon size to fit the data range and achieve regular hexagons without overlapping.

@@

```{r hex-size-img, out.width = "100%", out.height = "100%", fig.cap = "Hexagonal size parameter", fig.alt="Hexagonal size parameter", echo=FALSE}
knitr::include_graphics("figures/hex_size_img.png")
```

```{r spacing-img, out.width = "100%", out.height = "100%", fig.cap = "Horizontal and vertical spacing parameters", fig.alt="Horizontal and vertical spacing parameters", echo=FALSE}
knitr::include_graphics("figures/spacing_img.png")
```

### Number of bins along the x-axis ($b_1$)

In the process of computing hexagonal bin configurations, the first step involves determining the number of bins along the x-axis. This begins with the initialization of parameters such as the hexagonal size ($s$) and the buffer along the x-axis. Subsequently, the range of the first 2D embedding component ($r_1$) is calculated to ensure comprehensive coverage of the data. This range is then adjusted by incorporating the buffer amount, effectively expanding the boundary to accommodate potential outliers or edge cases. The default buffer amount along the x-axis is $\sqrt{3} * s * 1.5$. Next, the horizontal spacing ($h$) between hexagons is computed based on the hexagon size. Finally, the number of bins along the x-axis is computed based on the adjusted range ($r_1 + buffer_{x}$) and the horizontal spacing ($h$).

$$
h = \sqrt{3} * s
$${#eq-equation3}

$$
 b_1 = \frac{r_1 + buffer_{x}}{h}
$${#eq-equation4}

@@

### Number of bins along the y-axis ($b_2$)

In the process of computing hexagonal bin configurations, the second step involves determining the number of bins along the y-axis. This begins with the initialization of parameters such as the hexagonal size and the buffer along the y-axis. Subsequently, the range of the second 2D embedding component ($r_2$) is calculated to ensure comprehensive coverage of the data. This range is then adjusted by incorporating the buffer amount, effectively expanding the boundary to accommodate potential outliers or edge cases. The default buffer amount along the y-axis is $1.5 * s * 1.5$. Next, the vertical spacing ($v$) between hexagons is computed based on the hexagon size. Finally, the number of bins along the y-axis is computed based on the adjusted range ($r_2 + buffer_{y}$) and the vertical spacing ($v$).

$$
v = 1.5 * s
$${#eq-equation5}

$$
 b_2 = \frac{r_2 + buffer_{y}}{v}
$${#eq-equation6}

@@

Generating full hexagonal grid contains main four steps:

@@

The default $x_{start}$ and $y_{start}$ can be defined as:

$$
x_{start} = min(emb_1) - \frac{\sqrt{3} * s}{2}
$${#eq-equation7}

$$
y_{start} = min(emb_2) - \frac{1.5 * s}{2}
$${#eq-equation8}

```{r gen-fullcentroids, out.width = "100%", out.height = "100%", fig.cap = "Workflow of generating hexagonal bin centroids", fig.alt="Workflow of generating hexagonal bin centroids", echo=FALSE}
knitr::include_graphics("figures/workflow_all_centroids.png")
```

@@

$$
dx = \frac{x_{(2)} - x_{(1)}}{2}
$${#eq-equation9}

$$
dy = \frac{2*(y_{(2)} - y_{(1)}) * 1.15}{\sqrt{3}}
$${#eq-equation10}

```{r hex-coordinates, out.width = "100%", out.height = "100%", fig.cap = "Hexagon coordinates", fig.alt="Hexagon coordinates", echo=FALSE}
knitr::include_graphics("figures/hex_coordinates_img.png")
```

```{r gen-allcoordinates, out.width = "100%", out.height = "100%", fig.cap = "Workflow of generating hexagon coordinates", fig.alt="Workflow of generating hexagon coordinates", echo=FALSE}
knitr::include_graphics("figures/workflow_all_coordinates.png")
```

@@

3. Assign data into hexagons

@@

4. Compute standardized counts within each hexagon

@@

@@

```
<!--
#### Buffer size

@@

Overall, including a buffer when generating hexagonal bins helps to produce more accurate and robust binning results, particularly when dealing with real-world data that may have irregular distributions or boundary effects.
-->
### Construct the 2D model with different options

@@

### 2024-02-26 paper-quollr.tex
Add the hexagonal binning logic with images

\hypertarget{compute-hexagonal-bin-configurations}{%
\subsubsection{Compute hexagonal bin configurations}\label{compute-hexagonal-bin-configurations}}

@@

Before computing hexagonal bin configurations, we need to determine the range of data along the x and y axes to establish the boundary for hexagonal binning. Additionally, choosing an appropriate hexagon size (the radius of the outer circle) is essential. By default, the \texttt{calculate\_effective\_x\_bins()} and \texttt{calculate\_effective\_y\_bins()} functions use a hexagon size of 1.07457. However, users can adjust the hexagon size to fit the data range and achieve regular hexagons without overlapping.

@@

@@

@@

@@

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{4}
\tightlist
\item
  Assign data into hexagons
\end{enumerate}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{5}
\tightlist
\item
  Compute standardized counts
\end{enumerate}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{6}
\tightlist
\item
  Extract full grid info
\end{enumerate}

\begin{itemize}
\item
  Assign standardize counts for hex bins
\item
  Join with the hexagonal coordinates
\end{itemize}

@@

\includegraphics{paper-quollr_files/figure-latex/unnamed-chunk-16-1.pdf}

\hypertarget{buffer-size}{%
\paragraph{Buffer size}\label{buffer-size}}

When generating hexagonal bins in R, a buffer is often included to ensure that the data points are evenly distributed within the bins and to prevent edge effects. The buffer helps in two main ways:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \textbf{Preventing Edge Effects}: Without a buffer, the outermost data points might fall near the boundary of the hexagonal grid, leading to incomplete bins or uneven distribution of data. By adding a buffer, you create a margin around the outer edges of the grid, ensuring that all data points are fully enclosed within the bins.
\item
  \textbf{Ensuring Even Distribution}: The buffer allows for a smoother transition between adjacent bins. This helps in cases where data points are not perfectly aligned with the grid lines, ensuring that each data point is assigned to the nearest bin without bias towards any specific direction.
\end{enumerate}

Overall, including a buffer when generating hexagonal bins helps to produce more accurate and robust binning results, particularly when dealing with real-world data that may have irregular distributions or boundary effects.

\hypertarget{construct-the-2d-model-with-different-options}{%
\subsubsection{Construct the 2D model with different options}\label{construct-the-2d-model-with-different-options}}

@@

\hypertarget{shift-the-hexagonal-grid-origin}{%
\paragraph{Shift the hexagonal grid origin}\label{shift-the-hexagonal-grid-origin}}

If shift\_x happen to the positive direction of x it should input as a positive value, if not other way
If shift\_y happen to the positive direction of y it should input as a positive value, if not other way

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Assign shift along the x and y axis (limited the amount should less than the cell\_diameter)
\item
  Generate bounds with shift origin
\end{enumerate}

>>>>

\hypertarget{preprocessing}{%
\subsection{Preprocessing}\label{preprocessing}}

In the preprocessing stage, we aim to prepare our data to fit within the bounds required for regular hexagonal binning, ensuring effective visualization. To achieve this, we implement two key scaling steps. Firstly, we scale the first 2D embedding component to range between \(0\) and \(1\), ensuring that all data points fall within this normalized interval. Secondly, we scale the second 2D embedding component to range between \(-\frac{\sqrt{3}}{2}\) and \(\frac{\sqrt{3}}{2}\). This adjustment helps maintain the necessary symmetry and spacing required for regular hexagons.

\hypertarget{compute-hexagonal-bin-configurations}{%
\subsection{Compute hexagonal bin configurations}\label{compute-hexagonal-bin-configurations}}

@@

Before computing hexagonal bin configurations, we need to determine the range of data along the x and y axes to establish the boundary for hexagonal binning. Additionally, choosing an appropriate hexagon size (the radius of the outer circle) is essential. By default, the \texttt{calculate\_effective\_x\_bins()} and \texttt{calculate\_effective\_y\_bins()} functions use a hexagon size of \(0.2\). However, users can adjust the hexagon size to fit the data range and achieve regular hexagons without overlapping.

@@

\begin{figure}
\includegraphics[width=1\linewidth,height=1\textheight]{figures/hex_size_img} \caption{Hexagonal size parameter}\label{fig:hex-size-img}
\end{figure}

\begin{figure}
\includegraphics[width=1\linewidth,height=1\textheight]{figures/spacing_img} \caption{Horizontal and vertical spacing parameters}\label{fig:spacing-img}
\end{figure}

\hypertarget{number-of-bins-along-the-x-axis-b_1}{%
\subsubsection{\texorpdfstring{Number of bins along the x-axis (\(b_1\))}{Number of bins along the x-axis (b\_1)}}\label{number-of-bins-along-the-x-axis-b_1}}

In the process of computing hexagonal bin configurations, the first step involves determining the number of bins along the x-axis. This begins with the initialization of parameters such as the hexagonal size (\(s\)) and the buffer along the x-axis. Subsequently, the range of the first 2D embedding component (\(r_1\)) is calculated to ensure comprehensive coverage of the data. This range is then adjusted by incorporating the buffer amount, effectively expanding the boundary to accommodate potential outliers or edge cases. The default buffer amount along the x-axis is \(\sqrt{3} * s * 1.5\). Next, the horizontal spacing (\(h\)) between hexagons is computed based on the hexagon size. Finally, the number of bins along the x-axis is computed based on the adjusted range (\(r_1 + buffer_{x}\)) and the horizontal spacing (\(h\)).

\[
h = \sqrt{3} * s
\]\{\#eq-equation3\}

\[
 b_1 = \frac{r_1 + buffer_{x}}{h}
\]\{\#eq-equation4\}

@@

\hypertarget{number-of-bins-along-the-y-axis-b_2}{%
\subsubsection{\texorpdfstring{Number of bins along the y-axis (\(b_2\))}{Number of bins along the y-axis (b\_2)}}\label{number-of-bins-along-the-y-axis-b_2}}

In the process of computing hexagonal bin configurations, the second step involves determining the number of bins along the y-axis. This begins with the initialization of parameters such as the hexagonal size and the buffer along the y-axis. Subsequently, the range of the second 2D embedding component (\(r_2\)) is calculated to ensure comprehensive coverage of the data. This range is then adjusted by incorporating the buffer amount, effectively expanding the boundary to accommodate potential outliers or edge cases. The default buffer amount along the y-axis is \(1.5 * s * 1.5\). Next, the vertical spacing (\(v\)) between hexagons is computed based on the hexagon size. Finally, the number of bins along the y-axis is computed based on the adjusted range (\(r_2 + buffer_{y}\)) and the vertical spacing (\(v\)).

\[
v = 1.5 * s
\]\{\#eq-equation5\}

\[
 b_2 = \frac{r_2 + buffer_{y}}{v}
\]\{\#eq-equation6\}

@@

The default \(x_{start}\) and \(y_{start}\) can be defined as:

\[
x_{start} = min(emb_1) - \frac{\sqrt{3} * s}{2}
\]\{\#eq-equation7\}

\[
y_{start} = min(emb_2) - \frac{1.5 * s}{2}
\]\{\#eq-equation8\}

\begin{figure}
\includegraphics[width=1\linewidth,height=1\textheight]{figures/workflow_all_centroids} \caption{Workflow of generating hexagonal bin centroids}\label{fig:gen-fullcentroids}
\end{figure}

@@

\[
dx = \frac{x_{(2)} - x_{(1)}}{2}
\]\{\#eq-equation9\}

\[
dy = \frac{2*(y_{(2)} - y_{(1)}) * 1.15}{\sqrt{3}}
\]\{\#eq-equation10\}

\begin{figure}
\includegraphics[width=1\linewidth,height=1\textheight]{figures/hex_coordinates_img} \caption{Hexagon coordinates}\label{fig:hex-coordinates}
\end{figure}

\begin{figure}
\includegraphics[width=1\linewidth,height=1\textheight]{figures/workflow_all_coordinates} \caption{Workflow of generating hexagon coordinates}\label{fig:gen-allcoordinates}
\end{figure}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{2}
\tightlist
\item
  Assign data into hexagons
\end{enumerate}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{3}
\tightlist
\item
  Compute standardized counts within each hexagon
\end{enumerate}

@@

@@

\includegraphics{paper-quollr_files/figure-latex/unnamed-chunk-16-1.pdf}
\#\#\# Construct the 2D model with different options

@@
